Design Document: Monte Carlo Maximum Likelihood for Generalized Linear Mixed Models

### 2015-01-13 NewDesignDoc/DesignDocGLMM.tex
Updated design doc's first 5 sections

Then the Monte Carlo log likelihood approximation is
\begin{align}
l_{m}(\theta) &=\log \dfrac{1}{m} \sum_{k=1}^mf_\theta(y|u_k)  \dfrac{ f_\theta(u_k)   }{\tilde{f}(u_k)}\\
&= \log \dfrac{1}{m} \sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}. \label{eq:MCLAval}
\end{align}
When $\tilde{f}$ depends on $\theta$, the gradient vector of the MCLA with respect to $\theta$ is
\begin{align}
\nabla l_m(\theta)&= \dfrac{\sum_{k=1}^m    \left( \nabla \log f_\theta(u_k,y) - \nabla \log \tilde{f} (u_k)  \right) \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} }{\sum_{k=1}^m \left( \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right) }, \label{eq:MCLAgradient}
\end{align}
and the Hessian matrix of the MCLA is
\begin{align}
&\nabla^2 l_m(\theta)= \dfrac{   \sum_{k=1}^m \left[ \nabla^2 \log f_\theta(y,u_k)  -
   \nabla^2 \log \tilde{f}(u_k)   \right]  \dfrac{ f_\theta(y,u_k)}{\tilde{f}(u_k)}  }{\sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}}\\
&+ \dfrac{   \sum_{k=1}^m \left[ \nabla \log f_\theta(y,u_k)  -
   \nabla \log \tilde{f}(u_k) - \nabla l_m(\theta)   \right] \left[ \nabla \log f_\theta(y,u_k)  -
   \nabla \log \tilde{f}(u_k) -\nabla l_m(\theta)  \right]'  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}   }{\sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}}. \label{eq:MCLAhessian}
\end{align}
Details for these equations can be found in section \ref{sec:calcs}. Calculation for the gradient and hessian when $\tilde{f}$ is independent of $\theta$ can be found in section \ref{sec:calcsindep}.

Now any inference, such as maximum likelihood, can be performed on $l_m(\theta)$ and its derivatives.

\section{Model fitting function} 
This is be the primary function the user  uses. The user  specifies the response and the predictors using the R formula mini-language as interpreted by \texttt{model.matrix}. Let $n$ be the observed sample size and recall that $p$ is the number of fixed effects.  When the fixed effects are specified, the model matrix $X$  is created, where $X$ has dimensions $n \times p$.

The user  specifies the random effects in the same way as for the R function \texttt{reaster} in the R package
 \texttt{aster} \citep{aster-package}. That is, random effects are expressed using the R formula mini-language. Let $q_t$ be the number of random effects associated with variance component $\nu_t$. When the random effects are specified, a list of T model matrices are created and the $t$th model matrix $Z_t$ has dimensions $n \times q_t$.

The user also specifies the expontial family (details in section \ref{sec:fam}), the name of the data set, and the name of the variance components.

@@

%$\Box$ Most of the time (but not all the time), the random effects formula should be ``0+...'' If the user does not specify ``0+'' then I want to give a warning something like ``Did you mean to start your random effects formula with 0+? Most models require this, so please check if you meant to include it.''

@@

As a result of this command, \texttt{model.matrix}  makes $Z$ into a list of 6 design matrices. Since we have 3 distinct variance components, we want 3 design matrices. We take the design matrices that share a variance component and \texttt{cbind} them together. The result is 3 model matrices in the $Z$ list, one for each variance component. We  want to put them in order (1,2,3 according to \texttt{varcomps.equal}) so that the names for each model matrix  ("subject","year","repeat") are correctly assigned.  Also, the variance estimates  come out in this same order as well.

  After interpreting the model the user has specified, the next step is to find penalized quasi-likelihood (PQL) estimates.  The process of finding these estimates is detailed in section \ref{sec:pql}. The PQL estimates parameterize the importance sampling distribution $\tilde{f}(u_k)$ which generate the random effects.  More information on generated the random effects is in section \ref{sec:genRand}.

@@

Let $g$ be the canonical link function and $\mu$ be a vector of length $n$ such that
\begin{align}
g(\mu) = X \beta + Z U
\end{align}
The choice of the link function is related to the distribution of the data, $ f_\theta(y|u)$. If the data have a  Bernoulli distribution, the link is $\logit(\mu)$. If the data have a Poisson distribution, the link is $\log (\mu)$. Currently, \texttt{glmm} offers only these two choices for the family, but any exponential family could work and can be easily added later.

 For simplicity of future notation, let $\eta=g(\mu)=X \beta + Z U$. Let $c(\eta)$ denote the cumulant function such that the log of the data density can be written as
\begin{align}
Y' \eta - c(\eta) = \sum_i \left[ Y_i \eta_i - c(\eta_i)  \right]
\end{align}

@@

With this, calculating $c(\eta_i), c'(\eta_i),$ and  $c''(\eta_i)$ is as simple as: 
\begin{verbatim}
family.glmm$c(args)
family.glmm$cp(args)
family.glmm$cpp(args)
\end{verbatim}

For the Bernoulli distribution, we calculate these values (\texttt{c}, \texttt{cp}, and \texttt{cpp}) with careful computer arithmetic as follows. We also use the \texttt{log1p} function in R for $c(\eta_i)$. 
\begin{align}
c(\eta_i) &= \log(1+e^{\eta_i}) =
  \begin{cases}
    \log(1+e^{\eta_i}) & \text{if } \eta_i\leq 0,\\
    \eta_i+\log(e^{-\eta_i}+1) & \text{if } \eta_i >0,
  \end{cases}\\
c'(\eta_i)&=\dfrac{ e^{{\eta_{i}}}}{ 1+e^{{\eta_{i}}}} = \dfrac{1}{1+e^{-\eta_i}}\\
c''(\eta_i)&=   \dfrac{e^{{\eta_{i}}}}{  1+ e^{{\eta_{i}}} }  - \dfrac{e^{2{\eta_{i}}}}{    (  1+ e^{{\eta_{i}}})^2}  = \dfrac{1}{1+e^{-\eta_i}}\cdot \dfrac{1}{1+e^{\eta_i}}
\end{align}
 For poisson, these values (\texttt{c}, \texttt{cp}, and \texttt{cpp}) are
\begin{align}
c(\eta_i)&=e^{\eta_i}\\
c'(\eta_i)&=e^{{\eta_{i}}}\\
c''(\eta_i)&= e^{{\eta_{i}}}.
\end{align}

@@

Also, I have a function to check that the data are valid given the family type. If \texttt{family.glmm} is \texttt{bernoulli.glmm}, the data should be 0s or 1s. If \texttt{family.glmm} is \texttt{poisson.glmm}, then the data should be nonnegative integers.  If the data does not pass the check, the check returns an error message.

@@

\subsection{Equations}
Recall the log of the data density is
\begin{align}
\log f_\theta(y|u) &= Y' \eta +c(\eta) \\
&= \sum_{i} Y_{i} {\eta_{i}} - c({\eta_{i}})
\end{align}
where
\begin{align}
\eta=X\beta+ZU.
\end{align}
The derivative of this with respect to one component, $\eta_j$, is
\begin{align}
\dfrac{\partial}{\partial \eta_j} \log f_\theta(y|u)  = Y_j-c'(\eta_j).
\end{align}
The derivative of the component $\eta_j$ with respect to one of the fixed effect predictors, $\beta_{l}$, is
\begin{align}
\dfrac{\partial \eta_j}{\partial \beta_{l}} = X_{j{l}}
\end{align}

We'd like the derivative of the log of the data density with respect to $\beta$. This can be written using the chain rule as follows:
\begin{align}
\dfrac{\partial}{\partial \beta_{l}}  \log f_\theta(y|u) &= \dfrac{\partial \eta_j}{\partial \beta_{l}} \dfrac{\partial}{\partial \eta_j} \log f_\theta(y|u) \\
&= \left[ Y_j-c'(\eta_j) \right]  X_{j{l}}
\end{align}

The mixed partial derivative (with respect to $\beta_{l_1}$ and $\beta_{l_2}$) of the log data density can be written similarly:
\begin{align}
\dfrac{\partial^2}{\partial \beta_{l_1} \partial \beta_{l_2}}  \log f_\theta(y|u) &=\dfrac{\partial}{\partial \beta_{l_2}} \left( \left[ Y_j-c'(\eta_j) \right]  X_{j{l_1}} \right) \\
&= \dfrac{\partial \eta_j}{\partial \beta_{l_2}} \dfrac{\partial}{\partial \eta_j} \left( \left[ Y_j-c'(\eta_j) \right]  X_{j{l_1}} \right) \\
&= -X_{j{l_1}} X_{j{l_2}} c''(\eta_j) 
 \end{align}

Letting $c'(\eta)$ be a vector with components $c'(\eta_j)$, the first derivative of the log data density can be written in matrix form as:
\begin{align}
\dfrac{\partial}{\partial \beta}  \log f_\theta(y|u) = X' \left[ Y-  c'(\eta)  \right].
\end{align}

Letting $ c(\eta)$ be a diagonal matrix with components $c''(\eta_j)$, the second derivative of the log data density can be written in matrix form as:
\begin{align}
   \frac{\partial^2}{\partial \beta^2} \log f_\theta(y|u) =   X' [ -  c''(\eta) ] X
\end{align}

@@

\section{Random effect generation and calculations}
This section is focused on a vector of random effects $u$ with length $q$. We assume that $u \sim N(0,D)$ where $D$ depends on $\nu$. Section \ref{sec:getEk} details the  relationship between  $\nu$ and $D$ explicit. Section \ref{sec:genRand} describes the process of generating the random effects. Section \ref{sec:distRand} explains how to evaluate the distribution of the random effects and its first two derivatives.

%%getEk
\subsection{Constructing $D$ (\texttt{getEk})}\label{sec:getEk}
For this version of the package, we that assume that $D$ is diagonal.  Let $\nu$ be length $T$ with components $\nu_t$, and let $E_t,t=1,\ldots,T$ be diagonal matrices with indicators on the diagonal so that $\sum_{t=1}^T E_t = I$. That is, the diagonal entries  of $E_t$ indicate whether that random effect has $\nu_t$ as a variance component. Then $D=\sum_{t=1}^T\nu_t  E_t $.

If $q_t$ is the number of random effects associated with variance component $\nu_t$, then $q_t$ is also the number of nonzero entries in $E_t$ and $q=\sum_{t=1}^T q_t$ is the total number of random effects in the model.  
The \texttt{model.matrix} part of \texttt{glmm} has made  \texttt{Z} into a list with $T$ design matrices (one for each distinct variance component).  In order to create $E_t$, we need to go through and count the  number of columns ($q_t$) for each design matrix $t$ in the list.
 Then $q=\sum_{t=1}^T q_t$ is the total number of random effects in the model, and $D$  has $q$ rows (and columns). Thus, each $E_t$  has $q$ rows and columns as well.  Then we know that $E_1$ has $q_1$ ones on the diagonal, followed by zeros to fill the rest of the diagonal.  $E_2$ has $q_1$ zeros, then $q_2$ ones, then zeros for the rest of the diagonal.

\subsection{Generating random effects (genRand)}\label{sec:genRand}
The model fitting function \texttt{glmm}  calls the function that generates the random effects and  stores them for future calculations. Let $s$ be a vector of length $q$ that represents the random effects on the standard normal scale. That is, $u= D^{1/2}s$. Let $\sigma$ be a vector of length $T$ with components $\sqrt{\nu_t}$.  The  PQL step of \texttt{glmm} has resulted in PQL estimates for $\beta$, $\sigma$ and  $s$ denoted by $\beta^*$, $s^*$ and $\sigma^*$. Then let
\begin{align}
A^*=\sum_{t=1}^T E_t \sigma^*_t
\end{align}\
and
\begin{align}
 D^* =A^*A^*
\end{align}
be matrices based on PQL estimates. Also, we can ``unstandardize'' our PQL-based random effects:
\begin{align}
u^*&=A^*s^*.
\end{align}

 Recall that $m$ is the overall Monte Carlo sample size. That is, $\breve{u}_k,k=1\ldots,m$ are vectors of random effects are drawn from a multivariate standard normal. Let $p_1,p_2,p_3$ be proportions such that  $p_1+p_2+p_3=1$ and let $m_1=m p_1$, $m_2= m p_2$, $m_3=m p_3$. Let $D$ and $A=D^{1/2}$ be determined by the current iteration of \texttt{trust}. Construct $((Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})^{1/2} $ as described in section \ref{sec:sqrtmat}. Then we center and scale $\breve{u}_k$ as follows:
\begin{align}
u_k&= A\, \breve{u}_k, k=1,\ldots,m_1\\
u_k&= u^*+A^* \, \breve{u}_k, k=m_1+1,\ldots,m_1+m_2\\
u_k&= u^*+ ((Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})^{1/2} \,  \breve{u}_k, k=m_1+m_2+1,\ldots,m.
\end{align}
 The result or centering and scaling is that
\begin{enumerate}
\item $u_k,k=1,\ldots,m_1 \sim N(0,D)$
\item $u_k,k=m_1+1,\ldots,m_1+m_2 \sim N(u^*,D^*)$
\item $u_k,k=m_1+m2+1,\ldots,m \sim N(u^*,(Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})$.
\end{enumerate}
 Let $f(u_k|\mu,\Sigma)$ denote the pdf for $u_k \sim N(\mu,\Sigma)$. Then the importance sampling distribution $\tilde{f}$ is the following mixture distribution:
\begin{align}
 \tilde{f}(u) = p_1 \,  f(u \, | \, 0, D)+p_2 \, f(u \, | \, u^*, D^*)+p_3 \, f(u \, | \, u^*, (Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1}) \label{eq:ftwiddle}
\end{align}

\subsubsection{Rationale}
The first distribution $N(0,D)$ was chosen to ensure a Central Limit Theorem, which is proven in  section \ref{sec:CLT}.   The second distribution is chosen because it is centered at the PQL best guess of the random effect values $u^*$ and  has the PQL guess of the variance. The last distribution is centered at the PQL guess $u^*$ and has a variance based on the PQL penalized likelihood Hessian.  The idea of this last distribution is to generate random effects from a distribution whose Hessian matches that of the target distribution $f_\theta(y|u) f_\theta(u)$.

\subsubsection{Finding the square root of a variance matrix}\label{sec:sqrtmat}
The shifting and scaling of $\breve{u})_k$ relies on finding the square root of a variance matrix. In the first two distributions, the variance matrices $D$ and $D^*$ are diagonal, and therefore we simply take the root of the diagonal elements. For the third distribution, we can use eigendecomposition. Eigendecompositions take a little bit more time than Cholesky decompositions but are more stable. Let $\Sigma$ denote the nondiagonal variance matrix. Eigendecomposition of $\Sigma$ provides orthogonal matrix $O$ (containing the eigenvectors) and diagonal matrix $\Lambda$ (with diagonal entries $\lambda$ being the eigenvalues of $\Sigma$).  Then
\begin{align}
\Sigma^{1/2}= O \Lambda^{1/2} O'.
\end{align}
Finding $\Lambda^{1/2}$ is as easy as taking the root of the diagonal entries.  We know that $\Sigma^{1/2}$ is correct because
\begin{align}
\Sigma^{1/2} \Sigma^{1/2} &= O \Lambda^{1/2} O'O \Lambda^{1/2} O'\\ \nonumber
&= O \Lambda^{1/2}\Lambda^{1/2} O'\\\nonumber
&=O \Lambda O' \\\nonumber
&= \Sigma.
\end{align}

\subsubsection{NEED TO PUT THIS SOMEWHERE}

Recall that $D$ is a diagonal with dimension $q$. $D$ has $q_1$ entries being $\nu_1$, $q_2$ entries being $\nu_2$, ..., and $q_T$ entries being $\nu_T$. Since the distribution of $u_k=A \breve{u}_k$ is $N(0,D)$, we write its pdf as:
\begin{align}
f(u_k|\nu)=f(A\breve{u}_k|\nu)&= (2 \pi)^{q/2} |D|^{-1/2} \exp(-0.5 (A \breve{u}_k)' D^{-1} (A \breve{u}_k))\\
&=(2 \pi)^{q/2} |A|^{-1} \exp(-0.5 \breve{u}_k' A D^{-1} A \breve{u}_k)\\
&=(2 \pi)^{q/2} |A|^{-1} \exp(-0.5 \breve{u}_k' \breve{u}_k)
\end{align}
So that 
\begin{align}
\log f(A \breve{u}_k|\nu) &\propto - \log |A| -\dfrac{1}{2} \breve{u}_k' \breve{u}_k\\
&\propto -\dfrac{q_1}{2} \log \nu_1 +\ldots + -\dfrac{q_T}{2} \log \nu_T -\dfrac{1}{2} \breve{u}_k' \breve{u}_k
\end{align}
Then 
\begin{align}
\dfrac{\partial}{\partial \nu_t} \log f(A \breve{u}_k|\nu) = \dfrac{-q_t}{2 \nu_t}\\
\dfrac{\partial^2}{\partial \nu_t^2} \log f(A \breve{u}_k|\nu) = \dfrac{q_t}{2 \nu_t^2}
\end{align}

>>>>

Then the Monte Carlo log likelihood approximation is
\begin{align}
l_{m}(\theta) &=\log \dfrac{1}{m} \sum_{k=1}^mf_\theta(y|u_k)  \dfrac{ f_\theta(u_k)   }{\tilde{f}(u_k)}\\
&= \log \dfrac{1}{m} \sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}. \label{eq:MCLAval}
\end{align}
When $\tilde{f}$ does not depend on $\theta$, the gradient vector of the MCLA with respect to $\theta$ is
\begin{align}
\nabla l_m(\theta)&= \dfrac{\sum_{k=1}^m     \nabla \log f_\theta(u_k,y)  \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} }{\sum_{k=1}^m  \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)}  }, \label{eq:MCLAgradient}
\end{align}
and the Hessian matrix of the MCLA is
\begin{align}
\nabla^2 l_m(\theta)&= \dfrac{   \sum_{k=1}^m  \nabla^2 \log f_\theta(y,u_k)       \dfrac{ f_\theta(y,u_k)}{\tilde{f}(u_k)}  }{\sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}} \nonumber\\
&+ \dfrac{   \sum_{k=1}^m \left[ \nabla \log f_\theta(y,u_k)   - \nabla l_m(\theta)   \right] \left[ \nabla \log f_\theta(y,u_k)  - \nabla l_m(\theta)  \right]'  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}   }{\sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}}
\end{align}
Details for these derivatives can be found in section \ref{sec:calcsindep}. Calculation for the gradient and hessian when $\tilde{f}$ is dependent on $\theta$ can be found in section \ref{sec:calcs}.

Now any likelihood-based inference, such as maximum likelihood, can be performed on $l_m(\theta)$ and its derivatives.

\section{Model fitting function} 
The model fitting function is be the primary function the user  uses. The user  specifies the response and the predictors using the R formula mini-language as interpreted by \texttt{model.matrix}. Let $n$ be the observed sample size and recall that $p$ is the number of fixed effects.  As a result of the user specifying the fixed effects, the model fitting function creates matrix $X$, which has dimensions $n \times p$.

The user  specifies the random effects in the same way as the fixed effects. This is also how users of  \texttt{reaster} (of R package
 \texttt{aster} \citep{aster-package}) specify the random effects. That is, random effects are expressed using the R formula mini-language. Let $q_t$ be the number of random effects associated with variance component $\nu_t$. When the random effects are specified, a list of $T$ model matrices are created. The $t$th model matrix $Z_t$ has dimensions $n \times q_t$.

The user also specifies the expontial family (details in section \ref{sec:fam}), the name of the data set, and the names of the variance components.

@@

@@

Initially, \texttt{model.matrix}  makes $Z$ into a list of 6 design matrices. Since we have 3 distinct variance components, we want 3 design matrices. We take the design matrices that share a variance component and use \texttt{cbind} to bind them together. The result is 3 model matrices in the $Z$ list, one for each variance component. We  want to put them in order (1,2,3 according to \texttt{varcomps.equal}) so that the names for each model matrix  (``subject'',``year'',``repeat'') are correctly assigned.  The variance estimates  are eventually listed in the model summary in this same order as well.

  After interpreting the model the user has specified, the next step is to find penalized quasi-likelihood (PQL) estimates.  The process of finding these estimates is detailed in section \ref{sec:pql}. The PQL estimates parameterize the importance sampling distribution $\tilde{f}(u_k)$ which generate the random effects.  More information on generating the random effects is in section \ref{sec:genRand}.

@@

Let $g$ be the canonical link function and $\mu$ be a vector of length $n$ such that
\begin{align}
g(\mu) = X \beta + Z u
\end{align}
The choice of the link function is related to the distribution of the data, $ f_\theta(y|u)$. If the data have a  Bernoulli distribution, the link is $\logit(\mu)$. If the data have a Poisson distribution, the link is $\log (\mu)$. Currently, \texttt{glmm} offers only these two choices for the family, but any exponential family could work and can be easily added later.

 For simplicity of future notation, let $\eta=g(\mu)=X \beta + Z u$. Let $c(\eta)$ denote the cumulant function such that the log of the data density can be written as
\begin{align}
y' \eta - c(\eta) = \sum_i \left[ y_i \eta_i - c(\eta_i)  \right]
\end{align}

@@

With this family of functions, calculating $c(\eta_i), c'(\eta_i),$ and  $c''(\eta_i)$ is as simple as: 
\begin{verbatim}
family.glmm$cum(args)
family.glmm$cp(args)
family.glmm$cpp(args)
\end{verbatim}

For the Bernoulli distribution, we calculate these values (\texttt{cum}, \texttt{cp}, and \texttt{cpp}) as follows. In order to be careful with our computer arithmetic, we use the \texttt{log1p} function and the second set of equalities.  
\begin{align}
c(\eta_i) &= \log(1+e^{\eta_i}) =
  \begin{cases}
    \log(1+e^{\eta_i}) & \text{if } \eta_i\leq 0,\\
    \eta_i+\log(e^{-\eta_i}+1) & \text{if } \eta_i >0,
  \end{cases}\\
c'(\eta_i)&=\dfrac{ e^{{\eta_{i}}}}{ 1+e^{{\eta_{i}}}} = \dfrac{1}{1+e^{-\eta_i}}\\
c''(\eta_i)&=   \dfrac{e^{{\eta_{i}}}}{  1+ e^{{\eta_{i}}} }  - \dfrac{e^{2{\eta_{i}}}}{    (  1+ e^{{\eta_{i}}})^2}  = \dfrac{1}{1+e^{-\eta_i}}\cdot \dfrac{1}{1+e^{\eta_i}}
\end{align}
 For Poisson, these values (\texttt{cum}, \texttt{cp}, and \texttt{cpp}) are
\begin{align}
c(\eta_i)&=e^{\eta_i}\\
c'(\eta_i)&=e^{{\eta_{i}}}\\
c''(\eta_i)&= e^{{\eta_{i}}}.
\end{align}

@@

Also, this family of functions contains a check to ensure the data are valid given the family type. If \texttt{family.glmm} is \texttt{bernoulli.glmm}, the data should contain only $0$ and $1$. If \texttt{family.glmm} is \texttt{poisson.glmm}, then the data should be nonnegative.  If the data set does not pass the check, the check returns an error message.

@@

\subsection{Equations}
Recall the log of the data density is
\begin{align}
\log f_\theta(y|u) &= y' \eta +c(\eta) \\
&= \sum_{i} y_{i} {\eta_{i}} - c({\eta_{i}})
\end{align}
where
\begin{align}
\eta=X\beta+Zu.
\end{align}
The derivative of this with respect to one component, $\eta_j$, is
\begin{align}
\dfrac{\partial}{\partial \eta_j} \log f_\theta(y|u)  = y_j-c'(\eta_j).
\end{align}
The derivative of the component $\eta_j$ with respect to one of the fixed effect predictors, $\beta_{l}$, is
\begin{align}
\dfrac{\partial \eta_j}{\partial \beta_{l}} = X_{j{l}}
\end{align}

We'd like the derivative of the log of the data density with respect to $\beta$. The first step is using the chain rule as follows:
\begin{align}
\dfrac{\partial}{\partial \beta_{l}}  \log f_\theta(y|u) &= \dfrac{\partial \eta_j}{\partial \beta_{l}} \dfrac{\partial}{\partial \eta_j} \log f_\theta(y|u) \\
&= \left[ y_j-c'(\eta_j) \right]  X_{j{l}}
\end{align}

The mixed partial derivative (with respect to $\beta_{l_1}$ and $\beta_{l_2}$) of the log data density can be written similarly:
\begin{align}
\dfrac{\partial^2}{\partial \beta_{l_1} \partial \beta_{l_2}}  \log f_\theta(y|u) &=\dfrac{\partial}{\partial \beta_{l_2}} \left( \left[ y_j-c'(\eta_j) \right]  X_{j{l_1}} \right) \\
&= \dfrac{\partial \eta_j}{\partial \beta_{l_2}} \dfrac{\partial}{\partial \eta_j} \left( \left[ y_j-c'(\eta_j) \right]  X_{j{l_1}} \right) \\
&= -X_{j{l_1}} X_{j{l_2}} c''(\eta_j) 
 \end{align}

Letting $c'(\eta)$ be a vector with components $c'(\eta_j)$, the first derivative of the log data density can be written in matrix form as:
\begin{align}
\dfrac{\partial}{\partial \beta}  \log f_\theta(y|u) = X' \left[ y-  c'(\eta)  \right].
\end{align}

Letting $ c''(\eta)$ be a diagonal matrix with components $c''(\eta_j)$, the second derivative of the log data density can be written in matrix form as:
\begin{align}
   \frac{\partial^2}{\partial \beta^2} \log f_\theta(y|u) =   X' [ -  c''(\eta) ] X
\end{align}

@@

\section{Random effect generation and calculations}
This section is focused on a vector of random effects $u$ with length $q$.  Section \ref{sec:getEk} details the  relationship between  $\nu$ and $D$.
Section \ref{sec:ftwiddle} states the importance sampling distribution.
 Section \ref{sec:genRand} describes the process of generating the random effects. Section \ref{sec:distRand} explains how to evaluate the distribution of the random effects and its first two derivatives.

%%getEk
\subsection{Constructing $D$ (\texttt{getEk})}\label{sec:getEk}
Recall that, for this version of the package, we assume that $D$ is diagonal.  Let $\nu$ be length $T$ with components $\nu_t$, and let $E_t,t=1,\ldots,T$ be diagonal matrices with indicators on the diagonal so that $\sum_{t=1}^T E_t = I$. That is, the diagonal entries  of $E_t$ indicate whether that random effect has $\nu_t$ as its variance component. Then $D=\sum_{t=1}^T\nu_t  E_t $.

Recall $q_t$ is the number of random effects associated with variance component $\nu_t$. Then $q_t$ is also the number of nonzero entries in $E_t$ and $q=\sum_{t=1}^T q_t$ is the total number of random effects in the model.  
The model fitting function of \texttt{glmm} has made  $Z$ into a list with $T$ design matrices (one for each distinct variance component).  In order to create $E_t$, we need to go through and count the  number of columns ($q_t$) for each design matrix $t$ in the list.
 Then $q=\sum_{t=1}^T q_t$ is the total number of random effects in the model, and $D$  is $q \times q$. Thus, each $E_t$  is $q \times q$.   $E_1$ has $q_1$ ones on the diagonal, followed by zeros to fill the rest of the diagonal.  $E_2$ has $q_1$ zeros, then $q_2$ ones, then zeros for the rest of the diagonal. We continue this process to complete the construction of $E_t, t=1,\ldots, T$.

\subsection{Importance sampling distribution $\tilde{f}(u_k)$}\label{sec:ftwiddle}
 Let $s$ be a vector of length $q$ that represents the random effects on the standard normal scale. That is $s$ is defined such that, $u= D^{1/2}s$. Let $\sigma$ be a vector of length $T$ with components $\sqrt{\nu_t}$.  Prior to generating random effects, the model fitting function of \texttt{glmm} has performed PQL and recorded the PQL estimates for $\beta$, $\sigma$ and  $s$. The PQL estimates are denoted by $\beta^*$, $s^*$ and $\sigma^*$. Let
\begin{align}
A^*=\sum_{t=1}^T E_t \sigma^*_t
\end{align}\
and
\begin{align}
 D^* =A^*A^*
\end{align}
be matrices based on PQL estimates. We can ``unstandardize'' our PQL-based random effects:
\begin{align}
u^*&=A^*s^*.
\end{align}

Let $p_1,p_2,p_3$ be the proportions of the mixture distribution such that  $p_1+p_2+p_3=1$.  Let $f(u|\mu,\Sigma)$ denote the pdf for $u \sim N(\mu,\Sigma)$.  Let $\grave{f}(u)$ denote the pdf for $u \sim t_\zeta$, a $q$-dimensional multivariate standard $t$ with $\zeta$ degrees of freedom. Then the importance sampling distribution is:
\begin{align}
 \tilde{f}(u) = p_1  \grave{f}(u)+p_2  f(u \, | \, u^*, D^*)+p_3  f(u \, | \, u^*, (Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1}). \label{eq:ftwiddle}
\end{align}
Then
\begin{align}
 \log \tilde{f}(u) = \log \left( p_1  \grave{f}(u)+p_2  f(u \, | \, u^*, D^*)+p_3  f(u \, | \, u^*, (Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1}) \right)
\end{align}

The first component of the mixture distribution $t_\zeta$ is chosen to ensure  the gradient of the MCLA has a central limit theorem, which is proven in  section \ref{sec:CLT}.  The second component is chosen because it is centered at the PQL best guess of the random effect values $u^*$ and  has the PQL guess of the variance. The last component is centered at the PQL guess $u^*$ and has a variance based on the Hessian of the PQL penalized likelihood.  The idea of this last distribution is to generate random effects from a distribution whose Hessian matches that of the target distribution $f_\theta(y|u) f_\theta(u)$.

\subsection{Generating random effects (genRand)}\label{sec:genRand}

 Recall that $m$ is the overall Monte Carlo sample size.  Let $p_1,p_2,p_3$ be the proportions of the mixture distribution such that  $p_1+p_2+p_3=1$. Let $m_1=m p_1$, $m_2= m p_2$, $m_3=m p_3$. 
%Let $D$ and $A=D^{1/2}$ be determined by the current iteration of \texttt{trust}.
Then we draw 
\begin{align}
u_k &\sim t_\zeta ,k=1,\ldots,m_1\\
u_k &\sim N(u \, | \, u^*, D^*) ,k=m_1+1,\ldots,m_1+m_2\\
u_k &\sim N(u \, | \, u^*, (Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1}) ,k=m_1+m_2+1,\ldots,m.
\end{align}
Details for drawing from a nonstandard normal are in section \ref{sec:gennonstand}.

\subsubsection{Finding the square root of a variance matrix}\label{sec:sqrtmat}
In order to generate draws from a nonstandard normal distribution, we need to calculate the square root of a variance matrix. When the variance matrix is diagonal (for example, $D^*$), we simply take the root of the diagonal elements.  When the variance matrix is not diagonal (for example, in the second component of the mixture distribution), we can use eigendecomposition. Eigendecompositions take a little bit more time than Cholesky decompositions but are more stable.

Let $\Sigma$ denote the nondiagonal variance matrix. Eigendecomposition of $\Sigma$ provides orthogonal matrix $O$ (containing the eigenvectors) and diagonal matrix $\Lambda$ (with diagonal entries $\lambda$ being the eigenvalues of $\Sigma$).  Then
\begin{align}
\Sigma^{1/2}= O \Lambda^{1/2} O'.
\end{align}
Finding $\Lambda^{1/2}$ is as easy as taking the root of the diagonal entries.  We know that $\Sigma^{1/2}$ is correct because
\begin{align}
\Sigma^{1/2} \Sigma^{1/2} &= O \Lambda^{1/2} O'O \Lambda^{1/2} O'\\ \nonumber
&= O \Lambda^{1/2}\Lambda^{1/2} O'\\\nonumber
&=O \Lambda O' \\\nonumber
&= \Sigma.
\end{align}

\subsubsection{Generating from nonstandard normal distributions}\label{sec:gennonstand}
 Construct $A^*=(D^*)^{1/2}$ and $((Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})^{1/2} $ as described in section \ref{sec:sqrtmat}.

Let $\breve{u}_k,k=m_1+1\ldots,m$ be vectors of random effects that are drawn from the standard normal distribution. Then we center and scale $\breve{u}_k$ as follows:
\begin{align}
u_k&= u^*+A^* \, \breve{u}_k, k=m_1+1,\ldots,m_1+m_2\\
u_k&= u^*+ ((Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})^{1/2} \,  \breve{u}_k, k=m_1+m_2+1,\ldots,m.
\end{align}
 The result or centering and scaling is that
\begin{align}
 &u_k \sim N(u^*,D^*), \, k=m_1+1,\ldots,m_1+m_2\\
& u_k \sim N(u^*,(Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1}), \, k=m_1+m2+1,\ldots,m .
\end{align}
